Constraining gravitational wave amplitude birefringence with GWTC-3

### 2023-04-04 src/tex/ms.tex
gw170818 wording

As we will find in Sec.~\ref{sec:Results}, the effect of birefringence can be (partially) degenerate with other parameters besides source inclination.
Indeed, the frequency-dependent amplification or dampening of polarizations caused by $\kappa$ can sometimes be absorbed by changes in intrinsic parameters, like the spins, with concurrent adjustments to the source inclination and distance.
As a measure of the \ac{BH} spin magnitudes along the orbital angular momentum, we will use the effective spin parameter \cite{Damour:2001tu,Ajith:2009bn,Santamaria:2010yb}
% \begin{equation}
% \chi_{\rm eff} \equiv \frac{1}{M} \left(m_1 \left|\vec{\chi}_{1\parallel}\right| + m_2\left|\vec{\chi}_{2\parallel}\right|\right)
% \end{equation}
\begin{equation}
\chi_{\rm eff} \equiv \frac{1}{1+q} \left(\left|\vec{\chi}_{1\parallel}\right| + q\left|\vec{\chi}_{2\parallel}\right|\right)
\end{equation}
where $|\vec{\chi}_{i\parallel}|$ are the norms of the projections of the dimensionless spin vectors along the orbital angular momentum, and the  mass ratio is $q \equiv m_2/m_1 \leq 1$.
We will study the interplay between birefringence and precession by focusing on the posterior of the precessing spin parameter $\chi_p$, defined as \cite{Schmidt:2014iyl}
\begin{equation}
\chi_p \equiv \max\left\{ |\vec{\chi}_{1\perp}|,\, k |\vec{\chi}_{2\perp}|\right\} ,
\end{equation}
where $|\vec{\chi}_{i\perp}|$ are the norms of the projections of the dimensionless spin vectors onto the orbital plane at a reference time, and $k \equiv q\left(4 q +3\right) / \left(4 + 3q\right)$.
We chart this and other approximate degeneracies as part of the results presented in Sec.~\ref{sec:Results}.

@@

GW170818 produced the posterior most displaced from $\kappa=0$, when judged by $\mu_i/\sigma_i$ in Fig.~\ref{fig:violin_kappa}.
Figure \ref{fig:corner_GW170818} shows that this happens because birefringence opens up a region of parameter space with $\kappa>0$ for larger distances and smaller inclination angles than would be allowed in the GR case.
We can make sense of this by noting that an edge-on, nonprecessing source produces linearly-polarized waves, meaning that a smaller inclination leads the two circular polarizations to have similar amplitudes.
On the other hand, having $\kappa > 0$ enhances the left-handed modes during propagation, per Eq.~\eqref{eq:waveform_modification}.
The two effects can be balanced to yield the correct polarization ratio observed at the detector (predominantly left-handed, per the preference for $\cos\iota \approx -1$ in the \ac{GR} analysis), as long as the distance is also enhanced to yield the right amount of birefringence and overall signal power.
This is similar to the degeneracy mentioned in Sec.~\ref{sec:inclination}.
% Even though the degeneracy is broken by the frequency dependence to a large extent, it is still possible for getting an extra region with a nonzero $\kappa$.
% This is because there is still a significant likelihood in the region of $\kappa$ close to zero.
% We can see this effect in other events as well, but not as strongly as in GW170818.

It is difficult to unequivocally identify a specific feature of the GW170818 data that leads to this posterior structure.
However, it appears to be related to this event's support for precession, in conjunction with its uncommonly definite measurement of the polarization and spin angles \cite{Varma:2021csh}.
The relevance of precession is evident from the posterior on $\chi_p$ (Fig.~\ref{fig:corner_GW170818}, top right): allowing for $\kappa \neq 0$ leads to reduced support for precession.
We can understand this in reference to Fig.~\ref{fig:birefringence}: if only a short portion of the signal is observed, then the frequency-dependent signal enhancement or dampening due to birefringence can mimic the time-dependent amplitude modulation produced by a precession cycle.
Therefore, under these circumstances, similar morphologies can be obtained by setting $\chi_p > 0$ or $\kappa > 0$, as long as the distance and inclination can also be adjusted accordingly.

The fact that the birefringent analysis favors a high $\kappa$ rather than a high $\chi_p$ is likely a consequence of prior volume: many more configurations are available with long distances and high $\kappa$ than with short distances and small $|\kappa|$.
The preference for $\kappa > 0$ over $\kappa < 0$ (and, therefore, the lack of bimodality in the $\cos\iota$ and $\kappa$ posteriors), is likely related to both the definite measurement of left-handed polarizations and the specific phasing of the precession cycle.
The latter manifests as a precise constraint on the spin orientation and phase angles in the \ac{GR} analysis \cite{Varma:2021csh} (see also Appendix \ref{sec:corner_GW170818_appendix}).
The observed amplitude modulation (say, increasing vs decreasing towards the merger) likely determines the allowed sign of $\kappa$ for this event.
Indeed, the inferred polarization angle in the birefringent case is notably different from the \ac{GR} one.
We provide the full corner plot for all relevant parameters in Fig.~\ref{fig:corner_GW170818_appendix} in Appendix~\ref{sec:corner_GW170818_appendix}.

>>>>

As we will find in Sec.~\ref{sec:Results}, the effect of birefringence can be (partially) degenerate with other parameters besides source inclination.
Indeed, the frequency-dependent amplification or dampening of polarizations caused by $\kappa$ can sometimes be absorbed by changes in intrinsic parameters, like the spins, with concurrent adjustments to the source inclination and distance.
As a measure of the \ac{BH} spin magnitudes along the orbital angular momentum, we will use the effective spin parameter \cite{Damour:2001tu,Ajith:2009bn,Santamaria:2010yb}
% \begin{equation}
% \chi_{\rm eff} \equiv \frac{1}{M} \left(m_1 \left|\vec{\chi}_{1\parallel}\right| + m_2\left|\vec{\chi}_{2\parallel}\right|\right)
% \end{equation}
\begin{equation}
\chi_{\rm eff} \equiv \frac{1}{1+q} \left(\chi_{1\parallel} + q\chi_{2\parallel}\right)
\end{equation}
where $\chi_{i\parallel}$ are the norms of the projections of the dimensionless spin vectors along the orbital angular momentum, and the  mass ratio is $q \equiv m_2/m_1 \leq 1$.
We will study the interplay between birefringence and precession by focusing on the posterior of the precessing spin parameter $\chi_p$, defined as \cite{Schmidt:2014iyl}
\begin{equation}
\chi_p \equiv \max\left\{ \chi_{1\perp},\, k \chi_{2\perp}\right\} ,
\end{equation}
where $\chi_{i\perp}$ are the norms of the projections of the dimensionless spin vectors onto the orbital plane at a reference time, and $k \equiv q\left(4 q +3\right) / \left(4 + 3q\right)$.
We chart this and other approximate degeneracies as part of the results presented in Sec.~\ref{sec:Results}.

@@

GW170818 produced the posterior most displaced from $\kappa=0$, when judged by $\mu_i/\sigma_i$ in Fig.~\ref{fig:violin_kappa}.
Figure \ref{fig:corner_GW170818} shows that this happens because birefringence opens up a region of parameter space with $\kappa>0$ for larger distances and smaller inclination angles than would be allowed in the GR case.
We can make sense of this by noting that an edge-on, nonprecessing source produces linearly-polarized waves, meaning that a smaller inclination leads the two circular polarizations to have similar amplitudes.
On the other hand, having $\kappa > 0$ enhances the left-handed modes during propagation, per Eq.~\eqref{eq:waveform_modification}.
The two effects can be balanced to match the polarization ratio observed at the detector (predominantly left-handed, per the preference for $\cos\iota \approx -1$ in the \ac{GR} analysis), as long as the distance is also enhanced to yield the right amount of birefringence and overall signal power.
This is similar to the degeneracy mentioned in Sec.~\ref{sec:inclination}.
% Even though the degeneracy is broken by the frequency dependence to a large extent, it is still possible for getting an extra region with a nonzero $\kappa$.
% This is because there is still a significant likelihood in the region of $\kappa$ close to zero.
% We can see this effect in other events as well, but not as strongly as in GW170818.

It is difficult to unequivocally identify a specific feature of the GW170818 data that leads to this posterior structure.
However, it appears to be related to this event's support for precession, in conjunction with its uncommonly definite measurement of the polarization and spin angles \cite{Varma:2021csh} (see Appendix \ref{sec:corner_GW170818_appendix}).
The relevance of precession is evident from the posterior on $\chi_p$ (Fig.~\ref{fig:corner_GW170818}, top right): allowing for $\kappa \neq 0$ leads to reduced support for precession.
We can understand this in reference to Fig.~\ref{fig:birefringence}: if only a short portion of the signal is observed, then the frequency-dependent signal enhancement or dampening due to birefringence can mimic the time-dependent amplitude modulation produced by a precession cycle.
Therefore, under these circumstances, similar morphologies can be obtained by setting $\chi_p > 0$ or $\kappa > 0$, as long as the distance and inclination can also be adjusted accordingly.

The fact that the birefringent analysis favors a high $\kappa$ rather than a high $\chi_p$ can be explained a consequence of prior volume: many more configurations are available with long distances and high $\kappa$ than with short distances and small $|\kappa|$.
The preference for $\kappa > 0$ over $\kappa < 0$ (and, therefore, the lack of bimodality in the $\cos\iota$ and $\kappa$ posteriors), is likely related to both the definite measurement of left-handed polarizations and the specific phasing of the precession cycle.
The latter manifests as a precise constraint on the spin orientation and phase angles in the \ac{GR} analysis \cite{Varma:2021csh} (see also Appendix \ref{sec:corner_GW170818_appendix}).
The observed amplitude modulation (say, increasing vs decreasing towards the merger) likely determines the allowed sign of $\kappa$ for this event.
Indeed, the inference on the polarization angle in the birefringent case is notably narrower than in \ac{GR}.
We provide the full corner plot for all relevant parameters in Fig.~\ref{fig:corner_GW170818_appendix} in Appendix~\ref{sec:corner_GW170818_appendix}.
